---
title: "Homework 5 - Sorting"
author: "Nicolas Solomita"
date: "2/4/2020"
header-includes:
   - \usepackage{algorithm}
   - \usepackage{algorithmic}
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 1
Generalize the SELECT algorithm to deal also with repeated values and prove that it still belongs to $O(n)$.

We built the `three_way_partition` procedure inside the file `select.c` in which we had to implement the select algorithm. This method is similar to the `partition` method; however it divides the array in three sub-arrays: one with values $< k$, one with values $> k$ and the last one with values equal to k. In this way we can deal also with the problem of sorting arrays with repeated elements. To prove that it still belongs to $O(n)$, we can easily observe that besides some simple instructions that have constant time, the other ones are the same of the partition algorithm; thus the same total complexity.

### Exercise 3

In the algorithm SELECT, the input elements are divided into chunks of 5. Will the algorithm work in linear time if they are divided into chunks of 7? What about chunks of 3?

__Chunks of 7__
We want to divide the array into chunks of 7 elements, and then find the median $m_i$ of each chunk and the median $m$ of the medians.

- number of chunks: $\lceil\frac{n}{7}\rceil$
- number of $m_i \geq m$: $\lceil\frac{1}{2}\lceil\frac{n}{7}\rceil\rceil$
- number of chunks that have at least 4 elements greater than $m$: $\lceil\frac{1}{2}\lceil\frac{n}{7}\rceil\rceil-2$
- minimum number of elements that are greater than $m$: $4\lceil\frac{1}{2}\lceil\frac{n}{7}\rceil\rceil-2 \geq \frac{4}{14}n -8$

Therefore, the max size of S or G is $n-\Big(\frac{4}{14}n -8\Big) =  \frac{5}{7}n + 8$ 

$$
\begin{equation}
  T_S(n)=\begin{cases}
    \Theta(1) & \text{if $n=1$}\\
    T_S(\lceil\frac{n}{7}\rceil) + T_S\Big(\frac{5}{7}n + 8\Big) + \Theta(n) & \text{otherwise}
  \end{cases}
\end{equation}
$$
We solve this equation with the substitution method.

We will assume that $\forall m < n \Rightarrow T_S(m) \leq cm$.

We chose $cn$ as a representative of $O(n)$ and $c'n$ for $\Theta(n)$.

$$
\begin{align}
T_S(n) &= T_S(\lceil\frac{n}{7}\rceil) + T_S\Big(\frac{5}{7}n + 8\Big) + c'n \leq
\\ &\leq c(\frac{1}{7}n +1) + c (\frac{5}{7}n +8) + c'n =
\\ &= \frac{6}{7}cn + 9c +c'n
\end{align}
$$
We want that $T_S(n) \leq cn$, therefore:
$$
cn \geq \frac{6}{7}cn + 9c +c'n \Leftrightarrow c \geq \frac{7c'n}{n-63}
$$
Hence, $\exists c \space \forall n\geq63 \in \mathbb{N}\space s.t. T_S(n) \leq cn \in O(n)$

__Chunks of 3__

We want to divide the array into chunks of 3 elements, and then find the median $m_i$ of each chunk and the median $m$ of the medians.

Now the max size of S or G is $n-\Big(\frac{1}{3}n -4\Big) =  \frac{2}{3}n + 4$ 
$$
\begin{equation}
  T_S(n)=\begin{cases}
    \Theta(1) & \text{if $n=1$}\\
    T_S(\lceil\frac{n}{3}\rceil) + T_S\Big(\frac{2}{3}n + 4\Big) + \Theta(n) & \text{otherwise}
  \end{cases}
\end{equation}
$$
Where $\frac{2}{3}n + 4$ is the max size of S or G. We solve this equation with the substitution method.

We will assume that $\forall m < n \Rightarrow T_S(m) \leq cm$.

We chose $cn$ as a representative of $O(n)$ and $c'n$ for $\Theta(n)$.

$$
\begin{align}
T_S(n) &= T_S(\lceil\frac{n}{3}\rceil) + T_S\Big(\frac{2}{3}n + 4\Big) + c'n \leq
\\ &\leq c(\frac{1}{3}n +1) + c (\frac{2}{3}n +4) + c'n =
\\ &= cn + 5c +c'n
\end{align}
$$
We want that $T_S(n) \leq cn$, therefore:
$$
cn \geq cn + 5c +c'n \Leftrightarrow c \leq -\frac{c'n}{5}
$$
Hence, $\nexists c \space \forall n \in \mathbb{N}\space s.t. T_S(n) \leq cn \Rightarrow T_S(n) \notin O(n)$

### Exercise 4
Suppose that you have a “black-box” worst-case linear-time subroutine to get the position in $A$ of the value that would be in position $\frac{n}{2}$ if $A$ was sorted. Give a simple, linear-time algorithm that solves the selection problem for an arbitrary position $i$.

```{r, tidy=FALSE, eval=FALSE, highlight=FALSE }
def procedure(A, l, r, i):

  if r-l < 10:
    sort(A,l,r)
    return i
  endif

  median <- black_box(A, l, r)
  pos <- partition(A, l, r, median)

  if i==pos:
    return pos
  endif

  if i < pos:
    return procedure(A, l, pos-1, i)
  endif

  return procedure(A, pos+1, r, i)
enddef
```

### Exercise 5
Solve the following recursive equations by using both the recursion tree and the substitution method:

1. $T_1(n) = 2*T_1(\frac{n}{2}) + O(n)$ 
2. $T_2(n) = T_2(\lceil\frac{n}{2}\rceil) + T_2(\lfloor\frac{n}{2}\rfloor) + \Theta(1)$ 
3. $T_3(n) = 3*T_3(\frac{n}{2}) + O(n)$ 
4. $T_4(n) = 7*T_4(\frac{n}{2}) + \Theta(n^2)$  

__1 - Recursion Tree__

At the i-th step of the recursion tree there are $2^i$ recursive calls, each one taking time $c\Big(\frac{n}{2^i}\Big)$. 

The height of the tree is:
$$
i = log_2 n
$$
and the number of nodes in the last level is:
$$
2^{log_2 n} = n
$$
Now, we can compute the total complexity:
$$
\begin{align}
T_1(n) &\leq \sum_{i=0}^{log_2 n-1} 2^ic\Big(\frac{n}{2^i}\Big) + \Theta(n) =
\\ &= cn \sum_{i=0}^{log_2 n-1} 1 +  \Theta(n) =
\\ &= cn \space log_2 n +  \Theta(n)
\end{align}
$$
We can conclude that $T_1(n) \in O(n \space log_2n)$.

__1 - Substitution Method__

Let's prove the result just discovered.

We will assume that $\forall m < n \Rightarrow T_1(m) \leq cm \space log_2m$.

We chose $cn \space log_2 \space n$ as a representative of $O(n \space log_2n)$ and $c'n$ for $O(n)$.

$$
\begin{align}
T_1(n) &= 2T_1\Big(\frac{n}{2}\Big) + c'n \leq
\\ &\leq 2 c\frac{n}{2} \space log_2 \frac{n}{2} + c'n =
\\ &= cn \space log_2 n - cn + c'n
\end{align}
$$
We know that $c \leq c' \Leftrightarrow cn + c'n \leq 0$.

Hence, $\exists c \space \forall n \in \mathbb{N}\space s.t. T_1(n) \leq cn \space log_2n \in O(n \space log_2n)$

__2 - Recursion Tree__

The cost of each node of the tree is a constant $c$. But we have that every branch has different length. How many levels does every branch have? The left most is the longest with $h \leq log_2 2n$, instead the right most is the shortest with $h \geq log_2 \frac{n}{2}$.

Therefore, we have to divide our calculation:

__Right Branch__
$$
T_2(n) \geq \sum_{i=0}^{log_2 \frac{n}{2}} 2^ic = c \frac{2^{log_2 \frac{n}{2}+1}-1}{2-1} = c(2^{log_2 n}-1) = cn - c
$$
__Left Branch__
$$
T_2(n) \leq \sum_{i=0}^{log_2 2n} 2^ic = c \frac{2^{log_2 2n+1}-1}{2-1} = c(2^{log_2 n+2}-1) = 4cn - c
$$
We can conclude that $T2(n) \in O(n) \wedge T2(n) \in \Omega(n) \Leftrightarrow T2(n) \in \Theta(n)$.

__2 - Substitution Method__

Let's prove the result just discovered.

__Right Branch__

We will assume that $\forall m < n \Rightarrow T_2(m) \leq cm - d$.

We chose $cn -d$ as a representative of $O(n)$ and $1$ for $\Theta(1)$.

$$
\begin{align}
T_2(n) &= T_2(\lceil\frac{n}{2}\rceil) + T_2(\lfloor\frac{n}{2}\rfloor) + 1 \leq
\\ &\leq c\lceil\frac{n}{2}\rceil -d +c\lfloor\frac{n}{2}\rfloor -d +1  =
\\ &= cn -d-d+1
\end{align}
$$
We know that $d \leq 1 \Leftrightarrow -d +1 \leq 0$.

Hence, $\exists d \space \forall n \in \mathbb{N}\space s.t. T_2(n) \leq cn -d \in O(n)$

__Left Branch__

We will assume that $\forall m > n \Rightarrow T_2(m) \geq cm$.

We chose $cn$ as a representative of $\Omega(n)$ and $1$ for $\Theta(1)$.

$$
\begin{align}
T_2(n) &= T_2(\lceil\frac{n}{2}\rceil) + T_2(\lfloor\frac{n}{2}\rfloor) + 1 \geq
\\ &\geq c\lceil\frac{n}{2}\rceil +c\lfloor\frac{n}{2}\rfloor +1  \geq
\\ &\geq cn +1 \geq cn
\end{align}
$$
Hence, $\exists c \space \forall n \in \mathbb{N}\space s.t. T_2(n) \geq cn \in \Omega(n)$

__3 - Recursion Tree__

At the i-th step of the recursion tree there are $3^i$ recursive calls, each one taking time $c\Big(\frac{n}{2^i}\Big)$. 

The height of the tree is:
$$
i = log_2 n
$$
and the number of nodes in the last level is:
$$
3^{log_2 n} = n^{log_2 3}
$$
Now, we can compute the total complexity:
$$
\begin{align}
T_3(n) &\leq \sum_{i=0}^{log_2 n-1} 3^ic\Big(\frac{n}{2^i}\Big) + \Theta(n^{log_2 3}) =
\\ &= cn \sum_{i=0}^{log_2 n-1} \Big(\frac{3}{2}\Big)^i + \Theta(n^{log_2 3}) =
\\ &= cn \Big(\Big( \frac{3}{2}\Big)^{log_2 n} -1\Big) + \Theta(n^{log_2 3}) =
\\ &= 3^{log_2 n}c - cn + \Theta(n^{log_2 3})
\end{align}
$$
We can conclude that $T_3(n) \in O(n^{log_2 3})$.


__3 - Substitution Method__

Let's prove the result just discovered.

We will assume that $\forall m < n \Rightarrow T_3(m) \leq cm^{log_23}$.

We chose $cn^{log_23} - dn$ as a representative of $O(n^{log_2 3})$ and $c'n$ for $O(n)$.

$$
\begin{align}
T_3(n) &= 3T_3\Big(\frac{n}{2}\Big) + c'n \leq
\\ &\leq 3 c\Big(\frac{n}{2}\Big)^{log_2 3} - d\frac{n}{2} + c'n =
\\ &= cn^{log_2 3} -n(\frac{3}{2}d -c')
\end{align}
$$
We know that $\frac{3}{2}d \geq c' \Leftrightarrow \frac{3}{2}d -c' \geq 0$.

Hence, $\exists c \space \forall n \in \mathbb{N}\space s.t. T_3(n) \leq cn^{log_2 3} \in O(n^{log_2 3})$

__4 - Recursion Tree__

At the i-th step of the recursion tree there are $7^i$ recursive calls, each one taking time $c\Big(\frac{n}{2^i}\Big)^2$.

The height of the tree is:
$$
i = log_2 n
$$
and the number of nodes in the last level is:
$$
7^{log_2 n} = n^{log_2 7}
$$
Now, we can compute the total complexity:
$$
\begin{align}
T_4(n) &= \sum_{i=0}^{log_2 n-1} 7^ic\Big(\frac{n}{2^i}\Big)^2 + \Theta(n^{log_2 7}) =
\\ &= cn^2 \sum_{i=0}^{log_2 n-1} \Big(\frac{7}{4}\Big)^i + \Theta(n^{log_2 7}) =
\\ &= cn \Big(\Big( \frac{7}{4}\Big)^{log_2 n} -1\Big) + \Theta(n^{log_2 7}) =
\\ &= 7^{log_2 n}c - cn^2 + \Theta(n^{log_2 7})
\end{align}
$$
We can conclude that $T_4(n) \in \Theta(n^{log_2 7})$.

__4 - Substitution Method__

Let's prove the result just discovered.

We will assume that $\forall m < n \Rightarrow T_4(m) = cm^{log_27}$.

We chose $cn^{log_27} - dn^2$ as a representative of $\Theta(n^{log_2 7})$ and $c'n^2$ for $\Theta(n^2)$.

$$
\begin{align}
T_4(n) &=7T_4\Big(\frac{n}{2}\Big) + c'n^2 =
\\ &= 7 c\Big(\frac{n}{2}\Big)^{log_2 7} - 7d\Big(\frac{n}{2}\Big)^2 + c'n^2 =
\\ &= cn^{log_2 7} -n^2(\frac{7}{4}d - c')
\end{align}
$$
We know that $\frac{7}{4}d = c' \Leftrightarrow \frac{7}{4}d -c' = 0$.

Hence, $\exists c \space \forall n \in \mathbb{N}\space s.t. T_4(n) = cn^{log_2 7} \in \Theta(n^{log_2 7})$